Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2013-12-03 Dail_Mad_ext.tex
Incorporated comments from Richard plus formatted for Ecology

\documentclass[12pt]{article}
\renewcommand*\abstractname{Summary}
\usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines]{lineno} %,displaymath
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage{graphicx}
\usepackage[round,colon,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}
\usepackage{verbatim}
%\usepackage[pdftex,colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}
\usepackage[pdftex,hidelinks,pdfstartview={XYZ 1 1 1.5},bookmarks=true]{hyperref}
\usepackage{rotating}

\bibliographystyle{ecology}
%\hyphenpenalty=100000

\begin{document}

\title{Improved state-space models for inference about 
spatial and temporal variation in abundance from count data}

%\doublespacing
\author{Jeffrey A. Hostetler\\ %\footnote{Contact:
%     \url{HostetlerJ@si.edu}, +001-202-633-4205}\\
        \small Migratory Bird Center \\
        \small Smithsonian Conservation Biology Institute \\
        \small National Zoological Park, MRC 5503 \\
        \small Washington, DC 20013-7012\\
   \and Richard B. Chandler \footnote{Current address: Warnell School of
     Forestry and Natural Resources, University of Georgia,
     \url{rchandler@warnell.uga.edu}, +001-706-542-5815} \\
        \small USGS Patuxent Wildlife Research Center \\
        \small Laurel, MD 20708
}
\maketitle

\abstract{
\noindent
\begin{enumerate}
\item Models of population dynamics %play a central role in theoretical and
%applied ecology, where they 
are frequently used for purposes such as testing 
hypotheses about density dependence and predicting species' responses 
to future environmental change or conservation actions. Fitting models
of population dynamics to field data is challenging because most data
sets are characterized by missing data and observation error, which
can inflate estimates of process variation if ignored.  
%Failure to account for observation error in such models can result in 
%bias. Therefore, ecologists have increasingly relied on 
Recently, state-space models have been developed to deal with these problems
%these probdirectly model 
by directly modeling both the observation error and the ecological
process of interest.
\item Conventional state-space models, however, have
several important limitations: (1) they assume that random effects are
Gaussian distributed, which implies that abundance can be negative and
that false negative observation errors are equally likely as false positive
errors, %results in unrealistic predictions
%(e.g. negative values of abundance) and an implausible mode
%and observation error are Gaussian random effects
%the parameters are not identifiable
%in many common situations, 
(2) they do not admit spatial variation in
population dynamics, and (3) some of the parameters of the model are not
estimable. % they assert an implausible model for
%we have no clear interpretation of the
%observation error. 
We demonstrate how each of these problems can be
resolved using a class of hierarchical models proposed by
\citet[Biometrics]{dail_madsen:2011} 
%for spatially-replicated time series data 
that attributes observation error to imperfect detection.
\item We expand this class of models to accommodate classical growth
models (e.g. exponential and Ricker-logistic), zero-inflation,
and random effects. 
% such as observer-specific detection probabilities. 
We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods, as demonstrated by accompanying {\bf R} and {\bf JAGS} code.%. We provide code to fit these models using the \textbf{R} package
%\texttt{unmarked} and \textbf{JAGS} is also provided.
\item Results of a simulation study suggest that bias is %was 
negligible and coverage
nominal for the proposed model extensions. An analysis of data from
the North American Breeding Bird Survey highlights how these methods
can be readily applied to existing data, but it also suggests 
that precision will be low when direct information about detection
probability (such as is collected using distance sampling or
replicated counts) is lacking.
\end{enumerate}
}

@@

\textbf{Key words}: abundance, Dail and Madsen model, density-dependence,
detectability, Gompertz-logistic model, immigration, open population point 
count models, population dynamics, random observer effects, Ricker-logistic model, zero-inflated

@@

%Theoretical ecology requires models of population dynamics for testing
%hypotheses regarding spatial and temporal variation in
%abundance. For example, much theoretical work has focused on
Models of population dynamics are vital in both theoretical and
applied ecological research. For instance, %the have been used to assess
the importance and existence of phenomenon such as
density-dependent population regulation, population cycling, and
spatial synchrony have been studied by comparing data from natural
populations with theoretical models
\citep{may:1975,turchin:1990,bjornstad_etal:1999,royama:1977}. %,dennis_taper:1994
%and population models are required to evaluate associated hypotheses.
In applied contexts, population models are used for estimating extinction probabilities
\citep{schoener_spiller:1992,nadeem_lele:2011,hostetler_etal:2012} and for predicting the
effects of future environmental conditions or conservation actions on
population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.

@@

Recently developed state-space models %are a widely used approach for
have made it possible to %become the primary tool for 
study population dynamics while accounting for both process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_etal:2006}. Classical state-space
models are time series models in which the true state of the
system (e.g. population size during each year) is modeled as a hidden
process, and the observed data are modeled conditional on this process
and the observation error. One reason for the widespread adoption of
state-space models in ecology is that failure to account for 
%process variation and 
observation error can bias estimators of abundance and
population growth parameters. For instance, the strength of
density dependence will be over-estimated if observation error is
ignored \citep{link_nichols:1994,shenk_etal:1998}.

A simple state-space model can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$
and random process variation denoted $\eta_t$.
% Richard, is switching this from $\grave{o}$ to $\eta$ okay?  Andy thought
% the o with accent was weird and I think I agree.  $\nu$ seems standard,
% but we use it for a slightly different meaning (ES only) later.  
The growth model may be density-dependent, as in the case of the 
Ricker-logistic model, or it might be density-independent, such as when growth
is exponential.
%, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$ is the intrinsic rate of increase.
A generic model can be written: %The full model can now be written as:
\begin{subequations}
  \label{eq:ss1}
  \begin{align}
    N_1 &= X_1 \label{eq:ss1a} \\
N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \;
t=2,\hdots,T \label{eq:ss1b} \\
X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
t=1,\hdots,T. \label{eq:ss1c}
  \end{align}
\end{subequations}
%where $\eta_t$ is the random effect allowing for process
%variation unaccounted for by the deterministic model. 
In classical
state-space models, the two sets of random effects
are assumed to be i.i.d. Gaussian deviates:
$\eta_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. %Standard practice is to
%ignore 
The process variation associated with $N_1$ is often ignored, or
%, as indicated by Eq~\ref{eq:ss1a}. Alternatively, 
the population is assumed to be at equilibrium such that the $N_1$ can
be regarded as an outcome of the equilibrium distribution.

@@

In this paper, we focus on the model of \citet[henceforth the DM model]{dail_madsen:2011}
that resolves each of the problems outlined above. 
%simultaneous resolves each of the aforementioned problems with
%traditional state-space models, and is designed for simple count data.
%In the following section, we describe the DM model in its original
%form and explain how it resolves each of the deficiencies with
%standard state-space models. 
%In Section~\ref{sec:ext}, 
Our aim is to extend the
model to accommodate classical models of population growth and
to handle several features common to ecological
time series, such as excess zeros and unexplained random variation. 
%Specifically, we describe methods for accommodating
%excess zeros and nuisance variables such as random observer effects.
Both frequentist and Bayesian methods of inference are discussed, and
we evaluate the performance of the model using a simulation study by
analyzing data from the North American Breeding Bird Survey (BBS), one of
the most spatially and temporally extensive sets of count data on
vertebrate populations \citep{robbins_etal:1986}.
%code for fitting models is presented in the appendices.
%The Bayesian method
%is attractive in that it can accommodate prior information about
%detection parameters, which is available for many species, yet not
%collected as part of many existing monitoring programs.
%In Section~\ref{sec:app}, we evaluate the performance of the model
%extensions using a simulation study and by analyzing data from the
%North American Breeding Bird Survey (BBS), one of
%the most spatially and temporally extensive sets of count data on
%vertebrate populations \citep{robbins_etal:1986}. %The overarching aim
%of the paper is to provide ecologists with flexible and accessible means of
%addressing important questions related to the variation of abundance in
%space and time. 
%believe these extensions will increase the utility of the model for
%application to existing datasets such the Breeding Bird Survey.

\section{The Dail-Madsen Model}
\label{sec:dm}

@@

\subsection{The Original Formulation of the Model}

@@

\subsubsection{Initial abundance}

%Recall that 
Conventional state-space models assume that the
distribution for the initial time period %,
%$[N_1 | \bm{\theta}]$, 
is either the equilibrium
distribution, or has zero variance.
In contrast, \citet{dail_madsen:2011} proposed modeling $N_{i1}$
as either a Poisson or negative binomial random variable:
\begin{gather}
N_{i,1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
\text{or} \nonumber \\
N_{i,1} \sim \mathrm{NB}(\Lambda, \alpha)
\label{eq:N1}
\end{gather}
where $\Lambda_i$ is the expected abundance at site $i$ during
year 1.
The Poisson distribution assumes that the mean of $N_{i,1}$ is
equal to its variance, whereas the negative binomial distribution allows the
variance to be greater than the mean with the amount of
overdispersion determined by the parameter $\alpha$.
%In the Poisson case, the equivalent expression of
%$[N_1 | \bm{\theta}]$ (Eq.~\ref{eq:ss2a}) is
%$[N_{i1} | \Lambda] =\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
%\begin{equation}
%[N_{i1} | \Lambda] =
%\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
%  \label{eq:N1br}
%\end{equation}
% Maybe we don't need the Poisson distriubtion equation? - Jeff

@@

\subsubsection{Abundance in subsequent time periods}

The DM model assumes that abundance in time $t$ is a function of
abundance at time $t-1$, i.e. abundance at each site evolves as a
first order Markov process. %, although higher order processes are also possible.
\citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
however, in each case they modeled $N_t$ as the sum of two random variables:
$S_{i,t}$, the number of indivduals surviving from $t-1$ and not
emigrating; and $G_{i,t}$ the number of new individuals entering
the population. Their most general model was
\begin{equation}
\left.\begin{aligned}
S_{i,t}|N_{i,t-1} &\sim \mathrm{Bin}(N_{i,t-1}, \omega) \\
G_{i,t}|N_{i,t-1} &\sim \mathrm{Pois}(\gamma(N_{i,t-1})) \\
N_{i,t} &= S_{i,t}+G_{i,t}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
where $\omega$ is the apparent survival probability and $\gamma$
is the recruitment rate (which can depend on $N_{i,t-1}$).
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{i,t} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
depend on $N_{i,t-1}$, and which simulates a ``propagule rain'' of new
individuals; the autoregressive model, $G_{i,t} \sim
\mathrm{Pois}(\gamma(N_{i,t-1}))$, which
simulates geometric or density independent growth; and the
``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which assumes the
population is at equilibrium. Covariates of
$\omega$ and $\gamma$ can be easily accommodated, for example
using logit- and log-linear models respectively.

@@

\subsubsection{Observation process}

%Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model,
%i.e. the model for abundance dynamics. This state
%model has allowed us to replace the initial abundance
%distribution and transition models of conventional state-space models, with
%alternatives that (1) allow for modeling spatial variation in
%abundance, (2) respect the discrete nature of the counts, and
%(3) have parameters that are estimable.
%The one
%remaining task is to replace the observation model
%$X_t = N_t + \zeta_t$ (Eq.~\ref{eq:ss1c}) with one that can be
%more easily interpreted. 
The observation model adopted by \citet{dail_madsen:2011} is the same
binomial model proposed by \citet{royle:2004biom}: % assumes that individuals are missed due to imperfect
%detection. The simplest model for imperfect detection is
\begin{equation}
  X_{i,t} \sim \mathrm{Bin}(N_{i,t}, p)
  \label{eq:p1}
\end{equation}
where $p$ is the probability of detecting each individual. Variation
in detection probability can be modeled as a function of site-specific
or occasion-specific covariates using, for example, a logit-linear model.
%or equivalently,
%\begin{equation}
%[X_{it} | N_{it}, \bm{p}] = \binom{N_{it}}{X_{it}}
%\bm{p}^{X_{it}}(1-\bm{p})^{N_{it}-X_{it}}.
%  \label{eq:1}
%\end{equation}
%Under this model, $\bm{p}$ is interpreted as the detection
%probability parameter.

@@

\section{Model Extensions}
\label{sec:ext}

\subsection{Population growth models}

@@

Partitioning population growth into survival and recruitment
is ideal in terms of providing a mechanistic description of population
dynamics, but this is not always possible using simple count data,
especially when the sites are not closed with respect to immigration
and emigration. %Furthermore, a
%simpler model would have several merits: faster running time, fewer
%total model combinations, and possibly more realistic estimates.
If the mechanistic model must be sacrificed, we suggest replacing it
with classical population growth models that have a long history in
ecology. The exponential growth model is the simplest: 
$f(N_{i,t-1}) = N_{i,t-1}e^r$ where $r$ is the intrinsic
rate of increase. %Although demographic and environmental stochasticity
%in $r$ could be modeled as functions of covariates or random effects, a
%simpler way of 
To allow for demographic stochasticity, we can regard $N_{i,t}$ as a Poisson 
random variable, simplifying Eq.~\ref{eq:Nt} to:
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1}).
\label{eq:exp}
\end{equation}
This,
like the geometric-recruitment version of the model, is a variant on a simple
density-independent, exponential model of population growth.
Density-dependent versions of the model are also possible.  For
example:
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-N_{i,t-1}/K)))
\label{eq:rick}
\end{equation}
where $K$ is the stable equilibrium of the population and $r$ is
the instantaneous population growth rate at low population
densities, and both parameters are constrained to be positive. This is
a stochastic version of the Ricker-logistic population growth model
\citep{ricker:1954}. Another option is a modified Gompertz-logistic
density-dependent model \citep{hart_gotelli:2011}:
\begin{equation}
N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-\log(N_{i,t-1}+1)/\log(K+1))))
\label{eq:gomp}
\end{equation}
Here the interpretations of $r$ and $K$ are similar to in the
Ricker-logistic model. Because a single Poisson distribution controls
the distribution of $N_{i,t}$ in each of these models, the discrete 
convolution used by \citet{dail_madsen:2011} to construct the
likelihood is not required, speeding up processing time.

\subsection{Immigration models}

One limitation of the geometric-recruitment, exponential,
Ricker-logistic, and Gompertz-logistic versions of the DM models all
%share a common feature (or bug): once the 
is that they include no mechanism for a local population to recover
after extinction. %at a site reaches 0, it must remain at 0
%because all contributions to population growth are assumed to be
%local.  
However, these models can be generalized 
%We therefore generalized these models 
to include both internal and external (immigration) contributions
to population growth. For example, an exponential plus immigration
model is:
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1} + \iota)
  \label{eq:expimm2}
\end{equation}
where $\iota$ represents the average number of immigrants per year, and is
constrained to be
positive (this is equivalent to separate Poisson processes for
growth and immigration).  %This model is close to
%the constant-recruitment DM model (Eq.~\ref{eq:Nt}), with $\exp(r)$ instead
%of $\omega$ and $\iota$ instead of $\gamma$, except that the first
%process is Poisson distributed instead of binomial. 
The geometric-recruitment, Ricker-logistic,
and Gompertz-logistic models can be extended to allow for immigration in the
same way.   Deterministic versions of the Ricker-logistic + immigration and the 
Gompertz-logistic + immigration do reach stable equilibriums, but the equilibriums 
are not at $K$ and cannot be solved for analytically \citep{otto_day:2007}.  Therefore, we
refer to $K$ in these models as the semi-equilibrium abundance.
\begin{comment}
Similarly, the Ricker-logistic + immigration model can be represented as:
the
Gompertz-logistic + immigration model as: (13) and the geometric-recruitment +  immigration model as: (14)
\end{comment}

>>>>

\documentclass[12pt,english]{article}
%\renewcommand*\abstractname{Summary}
\usepackage[verbose,letterpaper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines]{lineno} %,displaymath
\linenumbers
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage{graphicx}
\usepackage[round,comma,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}
\usepackage{verbatim}
\usepackage{indentfirst}
\usepackage[format=plain,justification=raggedright,labelsep=period,font={stretch=1.9}]{caption}
%\usepackage[pdftex,colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}
\usepackage[pdftex,hidelinks,pdfstartview={XYZ 1 1 1.5}]{hyperref}
\usepackage{rotating}
\usepackage[compact]{titlesec}
\titleformat{\section}[display]{\normalfont\bfseries}{}{0em}{\centering}
\titleformat{\subsection}[display]{\normalfont\itshape}{}{0em}{\centering}
\bibliographystyle{ecology}
\setlength{\bibsep}{0.0pt}
\bibpunct{(}{)}{,}{a}{}{,}
%\hyphenpenalty=100000
\setlength{\jot}{-1.2ex}
\usepackage{etoolbox}
\AtBeginEnvironment{tabular}{\doublespacing}

\begin{document}
\renewcommand{\thefootnote}{\arabic{footnote}}
\title{Improved state-space models for inference about 
spatial and temporal variation in abundance from count data}

%\doublespacing
\author{Jeffrey A. Hostetler\thanks{Migratory Bird Center, Smithsonian 
Conservation Biology Institute, National Zoological Park, MRC 5503, 
Washington, DC 20013-7012}
   \and Richard B. Chandler\thanks{USGS Patuxent Wildlife Research Center 
        Laurel, MD 20708} \footnote{Current address: Warnell School of
     Forestry and Natural Resources, University of Georgia,
     \url{rchandler@warnell.uga.edu}, +001-706-542-5815}        
}
\date{} %remove date
\maketitle
\begin{spacing}{1.9}
\begin{flushleft}
\abstract{%\noindent 
Models of population dynamics %play a central role in theoretical and
%applied ecology, where they 
are frequently used for purposes such as testing 
hypotheses about density dependence and predicting species' responses 
to future environmental change or conservation actions. Fitting models
of population dynamics to field data is challenging because most data
sets are characterized by observation error, which
can inflate estimates of process variation if ignored.  
%Failure to account for observation error in such models can result in 
%bias. Therefore, ecologists have increasingly relied on 
Recently, state-space models have been developed to deal with this problem
%these probdirectly model 
by directly modeling both the observation error and the ecological
process of interest.
Conventional state-space models, however, have
several important limitations: (1) they assume that random effects are
Gaussian distributed, which implies that abundance can be negative and
that false positive observation errors are equally likely as false negative
errors; %results in unrealistic predictions
%(e.g. negative values of abundance) and an implausible mode
%and observation error are Gaussian random effects
%the parameters are not identifiable
%in many common situations, 
(2) they do not admit spatial variation in
population dynamics; and (3) some of the parameters of the model are not
estimable. % they assert an implausible model for
%we have no clear interpretation of the
%observation error. 
We demonstrate how each of these problems can be
resolved using a class of hierarchical models proposed by
\citet[Biometrics]{dail_madsen:2011} 
%for spatially-replicated time series data 
that attributes observation error to imperfect detection.
We expand this class of models to accommodate classical growth
models (e.g. exponential and Ricker-logistic), zero-inflation,
and random effects. 
% such as observer-specific detection probabilities. 
We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods, as demonstrated by accompanying {\bf R} and {\bf JAGS} code. %We provide code to fit these models using the \textbf{R} package
%\texttt{unmarked} and \textbf{JAGS} is also provided.
Results of a simulation study suggest that bias is %was 
negligible and coverage
nominal for the proposed model extensions. An analysis of data from
the North American Breeding Bird Survey highlights how these methods
can be readily applied to existing data, but it also suggests 
that precision will be low when direct information about detection
probability (such as is collected using distance sampling or
replicated counts) is lacking.
}

@@

\textit{Key words}: abundance, Dail and Madsen model, density-dependence,
detectability, Gompertz-logistic model, immigration, open population point 
count models, population dynamics, random observer effects, Ricker-logistic model, zero-inflated

@@

\section*{Introduction}
%Theoretical ecology requires models of population dynamics for testing
%hypotheses regarding spatial and temporal variation in
%abundance. For example, much theoretical work has focused on
Models of population dynamics are vital in both theoretical and
applied ecological research. For instance, %the have been used to assess
the importance and existence of phenomenon such as
density-dependent population regulation, population cycling, and
spatial synchrony have been studied by comparing data from natural
populations with theoretical models
\citep{may:1975,turchin:1990,bjornstad_etal:1999,royama:1977}. %,dennis_taper:1994
%and population models are required to evaluate associated hypotheses.
In applied contexts, population models are used for estimating extinction probabilities
\citep{schoener_spiller:1992,nadeem_lele:2011,hostetler_etal:2012} and for predicting the
effects of future environmental conditions or conservation actions on
population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.

@@

Recently developed state-space models %are a widely used approach for
have made it possible to %become the primary tool for 
study population dynamics while accounting for both process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_etal:2006}. Classical state-space
models are time series models in which the true state of the
system (e.g. population size during each year) is modeled as a hidden
process, and the observed data are modeled conditional on this process
and the observation error. One reason for the widespread adoption of
state-space models in ecology is that failure to account for 
%process variation and 
observation error can bias estimators of abundance and
population growth parameters. For instance, the strength of
density dependence will be overestimated if observation error is
ignored \citep{link_nichols:1994,shenk_etal:1998}.

A simple state-space model can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$
and random process variation denoted $\eta_t$.
% Richard, is switching this from $\grave{o}$ to $\eta$ okay?  Andy thought
% the o with accent was weird and I think I agree.  $\nu$ seems standard,
% but we use it for a slightly different meaning (ES only) later.  
The growth model may be density-dependent, as in the case of the 
Ricker-logistic model, or it might be density-independent, such as when growth
is exponential.
%, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$ is the intrinsic rate of increase.
A generic model can be written: %The full model can now be written as:
\begin{linenomath*}
\begin{gather}
  \label{eq:ss1}
  \begin{align}
    N_1 &= X_1 \nonumber \\
N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \;
t=2,\hdots,T  \\
X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
t=1,\hdots,T. \nonumber 
  \end{align}
\end{gather}
\end{linenomath*}
%where $\eta_t$ is the random effect allowing for process
%variation unaccounted for by the deterministic model. 
In classical
state-space models, the two sets of random effects
are assumed to be i.i.d. Gaussian deviates:
$\eta_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. %Standard practice is to
%ignore 
The process variation associated with $N_1$ is often ignored, or
%, as indicated by Eq~\ref{eq:ss1a}. Alternatively, 
the population is assumed to be at equilibrium such that the $N_1$ can
be regarded as an outcome of the equilibrium distribution.

@@

In this paper, we focus on the model of \citet[henceforth the DM model]{dail_madsen:2011}
that resolves each of the problems outlined above. 
%simultaneous resolves each of the aforementioned problems with
%traditional state-space models, and is designed for simple count data.
%In the following section, we describe the DM model in its original
%form and explain how it resolves each of the deficiencies with
%standard state-space models. 
%In Section~\ref{sec:ext}, 
Our aim is to extend the
model to accommodate classical models of population growth and
to handle several features common to ecological
time series, such as excess zeros and unexplained random variation. 
%Specifically, we describe methods for accommodating
%excess zeros and nuisance variables such as random observer effects.
Both frequentist and Bayesian methods of inference are discussed, and
we evaluate the performance of the model using a simulation study and by
analyzing data from the North American Breeding Bird Survey (BBS), one of
the most spatially and temporally extensive sets of count data on
vertebrate populations \citep{robbins_etal:1986}.
%code for fitting models is presented in the appendices.
%The Bayesian method
%is attractive in that it can accommodate prior information about
%detection parameters, which is available for many species, yet not
%collected as part of many existing monitoring programs.
%In Section~\ref{sec:app}, we evaluate the performance of the model
%extensions using a simulation study and by analyzing data from the
%North American Breeding Bird Survey (BBS), one of
%the most spatially and temporally extensive sets of count data on
%vertebrate populations \citep{robbins_etal:1986}. %The overarching aim
%of the paper is to provide ecologists with flexible and accessible means of
%addressing important questions related to the variation of abundance in
%space and time. 
%believe these extensions will increase the utility of the model for
%application to existing datasets such the Breeding Bird Survey.

\section*{The Dail-Madsen Model}
\label{sec:dm}

@@

%\subsection{The Original Formulation of the Model}

@@

\subsection*{Initial abundance}

%Recall that 
Conventional state-space models assume that the
distribution for the initial time period %,
%$[N_1 | \bm{\theta}]$, 
is either the equilibrium
distribution, or has zero variance.
In contrast, \citet{dail_madsen:2011} proposed modeling $N_{i1}$
as either a Poisson or negative binomial random variable:
\begin{linenomath*}
\begin{gather}
N_{i,1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
\text{or} \nonumber \\
N_{i,1} \sim \mathrm{NB}(\Lambda, \alpha)
\label{eq:N1}
\end{gather}
\end{linenomath*}
where $\Lambda_i$ is the expected abundance at site $i$ during
year 1.
The Poisson distribution assumes that the mean of $N_{i,1}$ is
equal to its variance, whereas the negative binomial distribution allows the
variance to be greater than the mean with the amount of
overdispersion determined by the parameter $\alpha$.
%In the Poisson case, the equivalent expression of
%$[N_1 | \bm{\theta}]$ (Eq.~\ref{eq:ss2a}) is
%$[N_{i1} | \Lambda] =\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
%\begin{equation}
%[N_{i1} | \Lambda] =
%\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
%  \label{eq:N1br}
%\end{equation}
% Maybe we don't need the Poisson distriubtion equation? - Jeff

@@

\subsection*{Abundance in subsequent time periods}

The DM model assumes that abundance in time $t$ is a function of
abundance at time $t-1$, i.e. abundance at each site evolves as a
first order Markov process. %, although higher order processes are also possible.
\citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
however, in each case they modeled $N_t$ as the sum of two random variables:
$S_{i,t}$, the number of indivduals surviving from $t-1$ and not
emigrating; and $G_{i,t}$ the number of new individuals entering
the population. Their most general model was
\begin{linenomath*}
\begin{equation}
\left.\begin{aligned}
S_{i,t}|N_{i,t-1} &\sim \mathrm{Bin}(N_{i,t-1}, \omega) \\
G_{i,t}|N_{i,t-1} &\sim \mathrm{Pois}(\gamma(N_{i,t-1})) \\
N_{i,t} &= S_{i,t}+G_{i,t}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
\end{linenomath*}
where $\omega$ is the apparent survival probability and $\gamma$
is the recruitment rate (which can depend on $N_{i,t-1}$).
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{i,t} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
depend on $N_{i,t-1}$, and which simulates a ``propagule rain'' of new
individuals; the autoregressive model, $G_{i,t} \sim
\mathrm{Pois}(\gamma(N_{i,t-1}))$, which
simulates geometric or density independent growth; and the
``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which assumes the
population is at equilibrium. Covariates of
$\omega$ and $\gamma$ can be easily accommodated, for example
using logit- and log-linear models respectively.

@@

\subsection*{Observation process}

%Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model,
%i.e. the model for abundance dynamics. This state
%model has allowed us to replace the initial abundance
%distribution and transition models of conventional state-space models, with
%alternatives that (1) allow for modeling spatial variation in
%abundance, (2) respect the discrete nature of the counts, and
%(3) have parameters that are estimable.
%The one
%remaining task is to replace the observation model
%$X_t = N_t + \zeta_t$ (Eq.~\ref{eq:ss1c}) with one that can be
%more easily interpreted. 
The observation model adopted by \citet{dail_madsen:2011} is the same
binomial model proposed by \citet{royle:2004biom}: % assumes that individuals are missed due to imperfect
%detection. The simplest model for imperfect detection is
\begin{linenomath*}
\begin{equation}
  X_{i,t} \sim \mathrm{Bin}(N_{i,t}, p)
  \label{eq:p1}
\end{equation}
\end{linenomath*}
where $p$ is the probability of detecting each individual. Variation
in detection probability can be modeled as a function of site-specific
or occasion-specific covariates using, for example, a logit-linear model.
%or equivalently,
%\begin{equation}
%[X_{it} | N_{it}, \bm{p}] = \binom{N_{it}}{X_{it}}
%\bm{p}^{X_{it}}(1-\bm{p})^{N_{it}-X_{it}}.
%  \label{eq:1}
%\end{equation}
%Under this model, $\bm{p}$ is interpreted as the detection
%probability parameter.

@@

\section*{Model Extensions}
\label{sec:ext}

\subsection*{Population growth models}

@@

Partitioning population growth into survival and recruitment
is ideal in terms of providing a mechanistic description of population
dynamics, but this is not always possible using simple count data,
especially when the sites are not closed with respect to immigration
and emigration. %Furthermore, a
%simpler model would have several merits: faster running time, fewer
%total model combinations, and possibly more realistic estimates.
When the mechanistic model is unrealistic, we suggest replacing it
with classical population growth models. %that have a long history in ecology.
The exponential growth model is the simplest: 
$N_{i,t} = N_{i,t-1}e^r$ where $r$ is the intrinsic
rate of increase. %Although demographic and environmental stochasticity
%in $r$ could be modeled as functions of covariates or random effects, a
%simpler way of 
To allow for demographic stochasticity, we can regard $N_{i,t}$ as a Poisson 
random variable, simplifying Eq.~\ref{eq:Nt} to:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1}).
\label{eq:exp}
\end{equation}
\end{linenomath*}

%This,
%like the geometric-recruitment version of the model, is a variant on a simple
%density-independent, exponential model of population growth.
Density-dependent versions of the model are also possible.  For
example:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-N_{i,t-1}/K)))
\label{eq:rick}
\end{equation}
\end{linenomath*}
where $K$ is the stable equilibrium of the population and $r$ is
the instantaneous population growth rate at low population
densities, and both parameters are constrained to be positive. This is
a stochastic version of the Ricker-logistic population growth model
\citep{ricker:1954}. Another option is a modified Gompertz-logistic
density-dependent model \citep{hart_gotelli:2011}:
\begin{linenomath*}
\begin{equation}
N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-\log(N_{i,t-1}+1)/\log(K+1))))
\label{eq:gomp}
\end{equation}
\end{linenomath*}
Here the interpretations of $r$ and $K$ are similar to in the
Ricker-logistic model. Because a single Poisson distribution controls
the distribution of $N_{i,t}$ in each of these models, the discrete 
convolution used by \citet{dail_madsen:2011} to construct the
likelihood is not required, speeding up processing time.

\subsection*{Immigration models}
